---
title: "IBS Genetic Distance Calculation Example"
author: "Your Name"
date: "`r Sys.Date()`"
output: md_document
---

## Introduction

This R Markdown document demonstrates how to calculate the IBS (Identity-By-State) genetic distance between genotypes using a sample genotype matrix. The distance calculation function is sourced directly from an external GitHub repository.

---

## Load Required Packages

We need the `reshape2` package for data transformation (specifically the `melt()` function) to process the genetic distance matrix into a long-form data frame.

```{r prep}
# Load necessary library
library(reshape2)
```

---

## Source Custom IBS Distance Function

The IBS distance function `ibs.dist` is sourced from the following GitHub repository:

- **Repository**: [https://github.com/ignaciojci/ibs-distance](https://github.com/ignaciojci/ibs-distance)

```{r load_ibs}
# Source the custom IBS distance calculation function
source("https://raw.githubusercontent.com/ignaciojci/ibs-distance/refs/heads/main/ibs-distance.R")
```

---

## Example Genotype Matrix

For demonstration, we create a small genotype matrix with 3 genotypes (Geno1, Geno2, Geno3) and 4 markers (Mark1 to Mark4). The values represent the allele counts per marker.

```{r matrix}
# Define a sample genotype matrix
m <- matrix(c(2,0,0,
              0,0,2,
              0,2,2,
              2,2,2),
            nrow=3,
            dimnames = list(paste0("Geno",1:3),
                            paste0("Mark",1:4)))

# Display the matrix
m
```

---

## Compute IBS Genetic Distance

We compute the genetic distance matrix using the `ibs.dist()` function with centering enabled.

```{r ibs}
# Calculate the IBS genetic distance matrix
gd <- ibs.dist(m, center=TRUE)

# Display the resulting distance matrix
gd
```

---

## Convert Distance Matrix to Long-Format Table

To make the pairwise distances easier to read and analyze, we reshape the distance matrix into a long-format data frame. Only the upper triangle of the symmetric distance matrix is used to avoid redundant pairs.

```{r melt}
# Convert distance matrix into long format for easier interpretation
gd_list <- melt(gd, varnames = c("LineA", "LineB"), value.name = "Genetic Distance")[upper.tri(gd), ]

# Display the formatted genetic distances
gd_list
```

---

## Conclusion

This document demonstrated how to:

1. Create a genotype matrix.
2. Calculate IBS genetic distances using a custom function.
3. Reshape and display the resulting distance data for interpretation.
